The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-08-22 paper2_JNLS/paper.tex
NM

Now it is possible to choose the scales $\ampscale, \lengthscale, \timescale$ for ease in working with a particular framework. We choose the following scales,
\begin{align}
\label{scales}
\ampscale = \pi^{1/2} \, \etastd \, , \qquad
\lengthscale = \frac{\lamfac \lam}{\pi} \, , \qquad
\timescale = \frac{\lamfac \lam}{\freqp \depth}
\end{align}
where $\lamfac$ is an integer to be chosen later. 
The explanation for these choices is as follows. First, we have chosen the characteristic amplitude, $\ampscale$, above to normalize the energy of the state-variable $u$ to unity, as will be demonstrated in Section \ref{tKdVSec}. Second, regarding $\lengthscale$, recall that $\lam$ is the characteristic wavelength corresponding to the peak forcing frequency $\freqp$ in the experiments. If only integer multiples of $\freqp$ were imposed (e.g.~lower frequencies were not present), then the forcing would produce waves that are periodic over lengthscale $\lam$. Since lower frequencies do exist, strict periodicity is not satisfied, but rather waves may be nearly periodic over the physical domain $\xi \in [-\lam,\lam]$. The approximation of near-periodicity becomes more accurate if integer multiples of $\lam$ are considered, i.e.~$\xi \in [-\lamfac \lam, \lamfac \lam]$. Thus, we have chosen $\lengthscale$ above so that the dimensionless domain of consideration is $\tilde{x} \in [-\pi, \pi]$ and periodic boundary conditions on $u$ can be  imposed on over this domain with accuracy that increases with $\lamfac$. Lastly, regarding $\timescale$, the most basic timescale in the experiments is simply $\freqp^{-1}$, i.e.~the period of waves passing a fixed reference point. Of course, the leading-order behavior in shallow water is simple translation of waves with uniform speed $c$. The KdV equation provides the first correction to this behavior and describes dynamics that evolve over longer timescales. Hence we have rescaled $\freqp^{-1}$ by the factor $N \lam/\depth \gg 1$, which provides a suitably long timescale and is in line with previous normalizations \cite{johnson1997modern}. The scales $\lengthscale$ and $\timescale$ change value across the ADC, which is important to note when comparing the theory against experimental measurements.

With the above choices, the dimensionless KdV takes the form
\begin{align}
\label{dimlessKdV}
&u_t + C_3 \drat^{-3/2} \, u u_x + C_2 \drat^{1/2} \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{3/2} \epsup \delup^{-1} \, , \quad
C_2 = \frac{\pi^3 \delup}{6 \lamfac^2} 
\end{align}
The constants $C_3$ and $C_2$ do {\em not} change value crossing the ADC and are given in terms of the dimensionless parameters
\begin{align}
&\epsup = \etastd / \dup
&&\mbox{\em upstream amplitude-to-depth ratio} \\
&\delup = \dup / \lamup
&&\mbox{\em upstream depth-to-wavelength ratio}
\end{align}
The reason for the subscripts $3$ and $2$ will become evident in the next section.

@@

The variable-depth KdV \eqref{dimlessKdV}, though not Hamiltonian throughout the entire domain, admits a Hamiltonian structure on each side of the ADC. Indeed, \eqref{dimlessKdV} can be expressed as
\begin{align}
\partial_t{u} = \sympJ \vard{\Hupdn}{u}
\end{align}
where $\sympJ = \pdi{}{x}$ is the symplectic operator and $\Ham = \Hupdn$ is the Hamiltonian, which will take different forms on either side of the ADC. It is convenient to decompose the Hamiltonian into a so-called cubic and quadratic component, given respectively by
\begin{align}
\label{H3H2}
\Hthree = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
\Htwo = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}
Then the Hamiltonian can be expressed as
\begin{equation}
\label{Hamiltonian}
\Hupdn = C_2 \dratupdn^{1/2} \, \Htwo - C_3 \dratupdn^{-3/2} \, \Hthree
\end{equation}
where $\drat = \dratupdn$ changes value across the ADC. More explicitly, substituting \eqref{dratpw}, the separate upstream and downstream Hamiltonians are
\begin{align}
&\Hup = C_2 \, \Htwo - C_3 \, \Hthree 						&& \text{for } t<0 \\
&\Hdn = C_2 \dratdn^{1/2} \, \Htwo - C_3 \dratdn^{-3/2} \, \Hthree	&& \text{for } t>0
\end{align}
The above formulation differs cosmetically from some previous ones \cite{abramov2003hamiltonian, bajars2013weakly, majda2019statistical}, which used the opposite sign in both $\sympJ$ and $\Ham$. With that sign convention, Majda et al. 2019 found that a {\em negative} inverse temperature is required to accurately describe the experimental observations \cite{majda2019statistical}. We have chosen the sign convention above so that a {\em positive} inverse temperature could be used and thus fit the theory into the most conventional form of statistical mechanics.

@@

We now introduce the truncated KdV (TKdV) system, which is the main focus of this study. Consider the Fourier series of the state variable
\begin{align}
&u(x,t) = \sum_{k=-\infty}^{\infty} \uhat_k e^{i k x} \, , \\
\label{uhat}
&\uhat_k = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x) e^{-i k x} \dx
\end{align}
Since $u$ is real valued, we have $\uhat_{-k} = \uhat_{k}^*$ and since momentum vanishes $\uhat_0 = 0$.
Next, consider the Galerkin truncation at wave number $\Lambda$
\begin{align}
\uL(x,t) = \Proj u =
\sum_{\abs{k} \le \Lambda} \uhat_k e^{i k x} \, , \qquad
\end{align}
where $\Proj$ is a projection operator and \eqref{uhat} still holds. Projecting the KdV equation \eqref{dimlessKdV} onto the finite dimensional space gives the truncated KdV equation (TKdV)
\begin{align}
\label{TKdV}
&\pd{\uL}{t} +  \frac{1}{2} C_3 \drat^{-3/2} \,\Proj (\uL)^2 + C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{3/2} \epsup \delup^{-1} \, , \quad
C_2 = \frac{\pi^3 \delup}{6 \lamfac^2} 
\end{align}
Note the additional projection operator in front of the quadratic term $\uL^2$, which removes the aliased modes of wavenumber larger than $\Lambda$. The constants are the same as before and have been repeated here for convenience.

@@

The Hamiltonian on either side of the ADC, $\Ham^{\pm}$, generates a corresponding mixed ensemble, or Gibbs measure given by 
\begin{align}
\Gupdn = C_{\thupdn} \exp(-\thupdn \Hupdn) \delta(\En - \pi)
\end{align}
Here $\theta = \thupdn$ is the inverse temperature and $C_{\theta}$ a constant (inverse of the partition function) that depends on $\theta$. Each measure has corresponding ensemble mean $\mean{}_{\pm}$.

@@

%----------------------------------------------------------------------------%
% Skewness formula
%----------------------------------------------------------------------------%
\subsection{Explicit formula for outgoing skewness}
An experimental observation is that the incoming skewness is small, $\meanup{H_3} \approx 0$. Inserting this approximation into the statistical matching condition \eqref{statmatch} and simplifying gives
\begin{equation}
\label{H3H2}
\frac{\meandn{H_3^+}} {\meandn{H_2^+} - \meanup{H_2^+}} = \frac{C_2}{C_3} \dratdn^2
\end{equation}
%where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
To convert to dimensional, i.e.~experimental, values we first note that
\begin{align}
\mean{H_3^+} = \frac{\pi}{3} \mean{u^3} = 
\frac{1}{3} \pi^{-1/2} \frac{\mean{\eta^3}}{\etastd^3} = \frac{1}{3} \pi^{-1/2} \skw(\eta)
\end{align}
Second, to convert the downstream surface slope, we note
\begin{align}
\pd{u}{\tilde{x}} = \frac{\lamfac \lamdn}{\pi^{3/2} \etastd} \pd{\eta}{\xi}
\end{align}
where slope with respect to $\xi$ is the same as with respect to dimensional $x$.
Note: I believe it is important here to use $\lamdn$ in reverting to dimensional variables since we are matching $\Hdn$.
Hence, we have
\begin{align}
&\mean{H_2^+} = \pi \mean{ \left( \pd{u}{\tilde{x}} \right)^2} = 
\left( \frac{\lamfac \lamdn}{\pi \etastd} \right)^2 \var(\eta_x)
\end{align}
%
Then algebra gives
\begin{equation}
\frac{\meandn{H_3}} {\meandn{H_2} - \meanup{H_2}} = 
\frac{\pi^{3/2} \, \etastd^2} {3 \lamfac^2 \lamdn^2} 
\left( \frac{\skwdn(\eta)} { \vardn(\eta_x) - \varup(\eta_x)} \right)
\end{equation}
Also we have the simplification
\begin{equation}
\frac{C_2}{C_3} = \frac{\pi^{3/2} \delup^2}{9 \lamfac^2 \epsup}
\end{equation}
Hence, \eqref{H3H2} gives the remarkably explicit formula
\begin{equation}
\frac{\skwdn(\eta)} {\vardn(\eta_x) - \varup(\eta_x)} = \frac{1}{3} \epsup^{-3} \dratdn^3 =
\frac{1}{3} \left( \frac{\ddn}{\etastd} \right)^3
\end{equation}
In particular, the ratio on the left, which can be measured in the experiments, is predicted to scale as the inverse cube of the wave amplitude and as the square of the depth ratio.
% Note: previously I had D^2 on the RHS. The power of 3 comes from using lamdn consistently in the nondimensionalization.

>>>>

Now it is possible to choose the scales $\ampscale, \lengthscale, \timescale$ for ease in working with a particular framework. We choose the following scales,
\begin{align}
\label{scales}
\ampscale = \pi^{1/2} \, \etastd \, , \qquad
\lengthscale = \frac{\lamfac \lam}{\pi} \, , \qquad
\timescale = \frac{\lamfac \lam}{\pi \freqp \depth}
\end{align}
where $\lamfac$ is an integer to be chosen later. 
The explanation for these choices is as follows. First, we have chosen the characteristic amplitude, $\ampscale$, above to normalize the energy of the state-variable $u$ to unity, as will be demonstrated in Section \ref{tKdVSec}. Second, regarding $\lengthscale$, recall that $\lam$ is the characteristic wavelength corresponding to the peak forcing frequency $\freqp$ in the experiments. If only integer multiples of $\freqp$ were imposed (e.g.~lower frequencies were not present), then the forcing would produce waves that are periodic over lengthscale $\lam$. Since lower frequencies do exist, strict periodicity is not satisfied, but rather waves may be nearly periodic over the physical domain $\xi \in [-\lam,\lam]$. The approximation of near-periodicity becomes more accurate if integer multiples of $\lam$ are considered, i.e.~$\xi \in [-\lamfac \lam, \lamfac \lam]$. Thus, we have chosen $\lengthscale$ above so that the dimensionless domain of consideration is $\tilde{x} \in [-\pi, \pi]$ and periodic boundary conditions on $u$ can be  imposed on over this domain with accuracy that increases with $\lamfac$. Lastly, regarding $\timescale$, the most basic timescale in the experiments is simply $\freqp^{-1}$, i.e.~the period of waves passing a fixed reference point. Of course, the leading-order behavior in shallow water is simple translation of waves with uniform speed $c$. The KdV equation provides the first correction to this behavior and describes dynamics that evolve over longer timescales. Hence we have rescaled $\freqp^{-1}$ by the factor $N \lam/\depth \gg 1$, which provides a suitably long timescale and is in line with previous normalizations \cite{johnson1997modern}. The scales $\lengthscale$ and $\timescale$ change value across the ADC, which is important to note when comparing the theory against experimental measurements.

With the above choices, the dimensionless KdV takes the form
\begin{align}
\label{dimlessKdV}
&u_t + C_3 \drat^{-3/2} \, u u_x + C_2 \drat^{1/2} \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{1/2} \epsup \delup^{-1} \, , \quad
C_2 = \frac{\pi^2 \delup}{6 \lamfac^2} 
\end{align}
The constants $C_3$ and $C_2$ do {\em not} change value crossing the ADC and are given in terms of the dimensionless parameters
\begin{align}
&\epsup = \etastd / \dup
&&\mbox{\em upstream amplitude-to-depth ratio} \\
&\delup = \dup / \lamup
&&\mbox{\em upstream depth-to-wavelength ratio}
\end{align}
The reason for the subscripts $3$ and $2$ will become evident in the next section.

@@

The variable-depth KdV \eqref{dimlessKdV}, though not Hamiltonian throughout the entire domain, admits a Hamiltonian structure on each side of the ADC. Indeed, \eqref{dimlessKdV} can be expressed as
\begin{align}
\partial_t{u} = \sympJ \vard{\Hupdn}{u}
\end{align}
where $\sympJ = \pdi{}{x}$ is the symplectic operator and $\Ham = \Hupdn$ is the Hamiltonian, which will take different forms on either side of the ADC. It is convenient to decompose the Hamiltonian into a so-called cubic and quadratic component, given respectively by
\begin{align}
\label{H3H2}
\Hthree = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
\Htwo = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}
Then the Hamiltonian can be expressed as
\begin{equation}
\label{Hamiltonian}
\Hupdn = C_2 \dratupdn^{1/2} \, \Htwo - C_3 \dratupdn^{-3/2} \, \Hthree
\end{equation}
where $\drat = \dratupdn$ changes value across the ADC. More explicitly, substituting \eqref{dratpw}, the separate upstream and downstream Hamiltonians are
\begin{align}
&\Hup = C_2 \, \Htwo - C_3 \, \Hthree 						&& \text{for } t<0 \\
&\Hdn = C_2 \dratdn^{1/2} \, \Htwo - C_3 \dratdn^{-3/2} \, \Hthree	&& \text{for } t>0
\end{align}
The above formulation differs cosmetically from some previous ones \cite{abramov2003hamiltonian, bajars2013weakly, majda2019statistical}, which used the opposite sign in both $\sympJ$ and $\Ham$. With that sign convention, Majda et al. 2019 found that a {\em negative} inverse temperature is required to accurately describe the experimental observations \cite{majda2019statistical}. We have chosen the sign convention above (which also happens to be consistent with Lax \cite{lax1975periodic}) so that a {\em positive} inverse temperature could be used and thus fit the theory into the most conventional form of statistical mechanics.

@@

We now introduce the truncated KdV (TKdV) system, which is the main focus of this study. Consider the Fourier series of the state variable
\begin{align}
&u(x,t) = \sum_{k=-\infty}^{\infty} \uhat_k e^{i k x} \, , \\
\label{uhat}
&\uhat_k = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x) e^{-i k x} \dx
\end{align}
Since $u$ is real valued, we have $\uhat_{-k} = \uhat_{k}^*$ and since momentum vanishes $\uhat_0 = 0$.
Next, consider the Galerkin truncation at wave number $\Lambda$
\begin{align}
\uL(x,t) = \Proj u =
\sum_{\abs{k} \le \Lambda} \uhat_k e^{i k x} \, , \qquad
\end{align}
where $\Proj$ is a projection operator and \eqref{uhat} still holds. Projecting the KdV equation \eqref{dimlessKdV} onto the finite dimensional space gives the truncated KdV equation (TKdV)
\begin{align}
\label{TKdV}
&\pd{\uL}{t} +  \frac{1}{2} C_3 \drat^{-3/2} \,\Proj (\uL)^2 + C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{1/2} \epsup \delup^{-1} \, , \quad
C_2 = \frac{\pi^2 \delup}{6 \lamfac^2} 
\end{align}
Note the additional projection operator in front of the quadratic term $\uL^2$, which removes the aliased modes of wavenumber larger than $\Lambda$. The constants $C_3$ and $C_2$ are the same as before and have been repeated here for convenience.

@@

The Hamiltonian on either side of the ADC, $\Ham^{\pm}$, generates a corresponding mixed ensemble, or Gibbs measure given by 
\begin{align}
\Gupdn = C_{\thupdn} \exp(-\thupdn \Hupdn) \delta(\En - 1)
\end{align}
Here $\theta = \thupdn$ is the inverse temperature and $C_{\theta}$ a constant (inverse of the partition function) that depends on $\theta$. Each measure has corresponding ensemble mean $\mean{}_{\pm}$.

@@

%----------------------------------------------------------------------------%
% Skewness formula
%----------------------------------------------------------------------------%
\subsection{Explicit formula for outgoing skewness}
An experimental observation is that the incoming skewness is small, $\meanup{H_3} \approx 0$. Inserting this approximation into the statistical matching condition \eqref{statmatch} and simplifying gives
\begin{equation}
\label{H3H2}
\frac{\meandn{\Hthree^+}} {\meandn{\Htwo^+} - \meanup{\Htwo^+}} = \frac{C_2}{C_3} \dratdn^2
\end{equation}
%where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
To convert to dimensional, i.e.~experimental, values we first note that
\begin{align}
\mean{\Hthree^+} = \frac{\pi}{3} \mean{u^3} = 
\frac{1}{3} \pi^{-1/2} \frac{\mean{\eta^3}}{\etastd^3} = \frac{1}{3} \pi^{-1/2} \skw(\eta)
\end{align}
Second, to convert the downstream surface slope, we note
\begin{align}
\pd{u}{\tilde{x}} = \frac{\lamfac \lamdn}{\pi^{3/2} \etastd} \pd{\eta}{\xi}
\end{align}
where slope with respect to $\xi$ is the same as with respect to dimensional $x$.
Note: I believe it is important here to use $\lamdn$ in reverting to dimensional variables since we are matching $\Hdn$.
Hence, we have
\begin{align}
&\mean{\Htwo^+} = \pi \mean{ \left( \pd{u}{\tilde{x}} \right)^2} = 
\left( \frac{\lamfac \lamdn}{\pi \etastd} \right)^2 \var(\eta_x)
\end{align}
%
Then algebra gives
\begin{equation}
\frac{\meandn{\Hthree}} {\meandn{\Htwo} - \meanup{\Htwo}} = 
\frac{\pi^{3/2} \, \etastd^2} {3 \lamfac^2 \lamdn^2} 
\left( \frac{\skwdn(\eta)} { \vardn(\eta_x) - \varup(\eta_x)} \right)
\end{equation}
Also we have the simplification
\begin{equation}
\frac{C_2}{C_3} = \frac{\pi^{3/2} \delup^2}{9 \lamfac^2 \epsup}
\end{equation}
Hence, \eqref{H3H2} gives the remarkably explicit formula
\begin{equation}
\frac{\skwdn(\eta)} {\vardn(\eta_x) - \varup(\eta_x)} = \frac{1}{3} \epsup^{-3} \dratdn^3 =
\frac{1}{3} \left( \frac{\ddn}{\etastd} \right)^3
\end{equation}
In particular, the ratio on the left, which can be measured in the experiments, is predicted to scale as the inverse cube of the wave amplitude and as the square of the depth ratio.
% Note: previously I had D^2 on the RHS. The power of 3 comes from using lamdn consistently in the nondimensionalization.
